A formal treatment of bidirectional typing

### 2023-10-12 tex/ESOP24/BiSig.tex
[ update ] Add the missing .bib
[ tex ] Revise until S6

\documentclass[orivec,envcountsame,envcountsect,envcountreset,oribibl]{llncs}

@@

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mostly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for syntax-directed simply typed languages: For every signature that specifies a bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a correct type synthesiser.
We propose a preprocessing step called `mode decoration', which helps the user to deal with missing type annotations.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
%\subfile{S7-formalisation}
\input{S8-future}
\appendix
%\input{appendix}

@@

\bibliography{../reference.bib}
%\IfFileExists{../reference.bib}{%
%}{\bibliography{ref}}

>>>>

\documentclass[orivec,envcountsame,envcountsect,oribibl]{llncs}

@@

%\let\theorem\relax
%\let\definition\relax
%\spnewtheorem{theorem}{Theorem}[section]{\bfseries}{}
%\spnewtheorem{definition}{Definition}[section]{\bfseries}{}

@@

\begin{abstract}
There has been much progress in designing bidirectional type systems and associated type synthesis algorithms, but mostly on a case-by-case basis.
This situation is in stark contrast to parsing, for which there have been general theories and widely applicable tools such as parser generators.
To remedy the situation, this paper develops a \emph{general} and \emph{formal} theory of bidirectional typing, and provides a \emph{verified} generator of \emph{proof-relevant} type synthesisers for simply typed languages: For every signature that specifies a mode-correct bidirectionally typed language, there exists a proof-relevant type synthesiser that for an input abstract syntax tree constructs a typing derivation if any, gives its refutation if not, or reports that the input does not have enough type annotations.
Soundness, completeness, and mode-correctness are studied universally for any simply typed language, which are sufficient conditions for deriving a type synthesiser.
We propose a preprocessing step called \emph{mode decoration}, which helps the user to deal with missing type annotations in an abstract syntax tree given by programmers.
The entire development is formalised in \Agda and can be further integrated with other language-formalisation frameworks.
\end{abstract}

\input{S1-intro}
\input{S2-outline}
\input{S3-defs.tex}
\input{S4-pre-synthesis}
\input{S5-synthesis}
\input{S6-example}
\subfile{S7-formalisation}
\input{S8-future}
\appendix
\input{appendix}

@@

\IfFileExists{reference}{
\bibliography{reference}
}{\bibliography{ref}}

### 2023-10-12 tex/ESOP24/S1-intro.tex
[ tex ] Revise until S6

Implementing a type synthesiser correctly, especially for advanced type systems, is non-trivial.
For example, proof assistants in use including \Coq and \Agda have many bugs in their type synthesis such as absurdity, supposedly equal terms which were not normalised to the same canonical term, and so on.
\LT{some GitHub issues \cite{Felicissimo2023}}
The gap between the design of type synthesis and its implementation cannot be ignored.

To mitigate the above problem, we demonstrate a general and formal treatment of bidirectional typing, providing an out-of-the-box \emph{verified generator} of \emph{proof-relevant} type synthesisers developed with the proof assistant \Agda:

@@

Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
Their treatment deviates from the common formulation: 
Traditionally, a type synthesis algorithm is presented as \emph{algorithmic system} relations such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then accompanied with soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if there are enough annotations.
By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial.
In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
It yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Hence, both algorithmic soundness and completeness in original form are implied.
On the other hand, the proof of this statement in type theory is a \emph{proof-relevant} type synthesiser, because it does not only return yes/no but construct a proof---either a typing derivation or its refutation. 
\LT{emphasise that our proof is an actual program}
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

@@

To make the notion of mode-correctness precise, we give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, which extends the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
Such a general definition allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct bidirectional binding signatures.
The latter theorem effectively amounts to a verified proof-relevant bidirectional-type-synthesiser generator for all syntax-directed bidirectional simple type systems that are mode-correct.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filing the gap between theory and implementation.

>>>>

To mitigate the above problem, we develop with the proof assistant \Agda a general and formal treatment of bidirectional typing and provide a \emph{verified} generator of \emph{proof-relevant} type synthesisers taking a language specification and a (scoped-checked) raw term as input.

@@

Our work adapts proof-relevant (bidirectional) type synthesis illustrated by \citet{Wadler2022}.
Their treatment deviates from the usual formulation: 
a type synthesis algorithm is traditionally presented as \emph{algorithmic rules} such as $\Gamma |- \isTerm{t} \syn A \mapsto \isTerm{t}'$, denoting that annotations can be added to $\isTerm{t}$ in the surface language to produce $\isTerm{t}'$ of type $A$ in the core language.
Such an algorithm is then followed by soundness and completeness assertions such that the algorithm correctly synthesises the type of a raw term and every typable term can be synthesised if sufficiently annotated.
By contrast, \citeauthor{Wadler2022} exploit the computational and logical nature of Martin-L\"of type theory and formulate \emph{algorithmic soundness, completeness, and decidability in one go}.
%(although they do not emphasise the difference of their approach from the traditional one)

Recall that the law of excluded middle $P + \neg P$ does not hold as an axiom for every $P$ in type theory, and we say $P$~is logically \emph{decidable} if the law holds for~$P$.
For example, a proof of \emph{decidable equality}, i.e. $\forall x, y.~(x = y) + (x \neq y)$, decides whether $x$~and~$y$ are equal and accordingly gives an identity proof or a refutation explicitly; such a decidability proof may or may not be possible depending on the domain of $x$~and~$y$, and is non-trivial in general.
In type theory all proofs as programs terminate, so logical decidability implies algorithmic decidability.
Further, suppose a proof of the following statement:
\begin{quote}
  `For a context $\Gamma$ and a raw term $t$, either a typing derivation of\, $\Gamma |- t : A$ exists for some type~$A$ or any derivation of\, $\Gamma |- t : A$ for some type~$A$ leads to a contradiction'
\end{quote}
or rephrased succinctly as 
\begin{quote}
  `It is \underline{\emph{decidable}} for any $\Gamma$~and~$t$ whether $\Gamma |- t : A$ is derivable for some~$A$'.
\end{quote}
Computationally, the proof yields either a typing derivation for the given raw term~$t$ or a negation proof that such a derivation is impossible where the former case is algorithmic soundness and the latter algorithmic completeness in contrapositive form.
Hence, both algorithmic soundness and completeness in original form are implied.
Hence, the formal proof of the statement as a program is a verified \emph{proof-relevant} type synthesiser, because it does not only return yes/no but also constructs a proof. 
%For a type synthesiser being proof-relevant, all of its intermediate programs have to be proof-relevant, including bidirectional type synthesiser and its mode decorator.

@@

To make the notion of mode-correctness precise, we first give a general definition of bidirectional simple type systems, called \emph{bidirectional binding signature}, extending the sorted version of \varcitet{Aczel1978}{'s binding signature} with bidirectionality.
A general definition of typed languages allows us to define mode-correctness and to investigate its consequences rigorously, including the uniqueness of synthesised types and the decidability of bidirectional type synthesis for all mode-correct signatures.
The latter theorem effectively amounts to a generator of proof-relevant bidirectional type synthesiser for all syntax-directed bidirectional simple type systems that are mode-correct.

@@

%
In short, we develop a general and formal theory of bidirectional type synthesis for simply typed languages, including 
\begin{enumerate}
  \item general definitions for bidirectional type systems and mode-correctness,
  \item the concept of mode decoration and the comparison between our completeness and \varcitet{Dunfield2021}{'s annotatability}, 
    \Josh{generalisation mode decoration?}
  \item rigorously proven consequences of mode-correctness---the uniqueness of synthesised types and the decidability of bidirectional type synthesis, which amounts to
  \item a fully verified generator for proof-relevant type-synthesisers.
\end{enumerate}
Our theory was first developed in \Agda and then translated to the mathematical vernacular, filling the gap between our informal presentation and its verified implementation.

### 2023-10-12 tex/ESOP24/S3-defs.tex
[ tex ] Revise until S6

The definitions are formulated in two steps:
\begin{enumerate*}
  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
  } of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{enumerate*}
%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

A simply typed language specifies
\begin{enumerate*}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{enumerate*}
Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.

@@

\begin{definition}
\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a language $(\Sigma, \Omega)$ with free variables in~$V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
    \\
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
    \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
  \begin{enumerate*}
    \item variables in $V$,
    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
  \end{enumerate*}
  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{definition}

@@

The definition of typing derivations is a bit more involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:

@@

\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  \begin{figure}
    \centering
    \small
    \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional system $(\Sigma, \Omega)$} 
    \begin{mathpar}
      \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
      \and
      \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
      \\
      \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
        \cdots \\
      \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
      {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
      \\ \text{for $\biop$ in $\Omega$}
    \end{mathpar}
    \caption{Bidirectional typing derivations}
    \label{fig:bidirectional-typing-derivations}
  \end{figure}
  \begin{figure}
    \centering
    \small
    \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for $(\Sigma, \Omega)$}
    \begin{mathpar}
      \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
      \and
      \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
      \and
      \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
      \and
      \inferrule{V, \vars{x}_1 |-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \\ \cdots \\ V, \vars{x}_{n} |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
      {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
      \and \text{for $\biop$}
    \end{mathpar}
    \caption{Mode derivations}
    \label{fig:mode-derivations}
  \end{figure}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
\end{definition}

Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.

>>>>

The definitions are formulated in two steps:
\begin{inlineenum}
  \item first we introduce a notion of arity and a notion of signature which includes a set\footnote{%
    Even though our theory is developed in Martin-L\"of type theory, the term `set' is used instead of `type' to avoid the obvious confusion. 
    Also, as we assume \AxiomK, all types are legitimately sets in the sense of homotopy type theory.
  } of operation symbols and an assignment of arities to symbols;
\item then, given a signature, we define raw terms and typing derivations inductively by primitive rules such as $\Rule{Var}$ and a rule schema for constructs $\tmOp_o$ indexed by an operation symbol~$o$.
\end{inlineenum}
%All these definitions are inductive as usual but include a \emph{rule schema} giving rise to as many rules as there are operation symbols in a signature.
As we move from simple types to bidirectional typing, the notion of arity, initially as the number of arguments of an operation, is enriched to incorporate an extension context for variable binding and the mode for the direction of type information flow.

@@

A simply typed language specifies
\begin{inlineenum}
  \item a family of sets of raw terms $\isTerm{t}$ indexed by a list~$V$ of variables (that are currently in scope) where each construct is allowed to bind some variables like $\Rule{Abs}$ and to take multiple arguments like $\Rule{App}$;
  \item a family of sets of typing derivations indexed by a typing context~$\Gamma$, a raw term~$\isTerm{t}$, and a type~$A$.
\end{inlineenum}
Therefore, to specify a term construct, we enrich the notion of arity with some sort for typing and extension context for variable binding.

@@

\begin{definition}
\begin{figure}
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}}{$\isTerm{t}$ is a raw term for a language $(\Sigma, \Omega)$ with free variables in~$V$}
  \begin{mathpar}
    \inferrule{\isTerm{x} \in V}{V |-_{\Sigma, \Omega} \isTerm{x}}\,\Rule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}}{V |-_{\Sigma, \Omega} \isTerm{t \annotate A}}\,\Rule{Anno}
    \\
    \inferrule{V, \vars{x}_\isTerm{1} |-_{\Sigma, \Omega} \isTerm{t_1} \\ \cdots \\ V, \vars{x}_\isTerm{n} |-_{\Sigma, \Omega} \isTerm{t_n} } {V |-_{\Sigma, \Omega} \tmOpts }\,\Rule{Op} \and
    \text{for $\bop$ in $\Omega$}
  \end{mathpar}
  \caption{Raw terms}
  \label{fig:raw-terms}
\end{figure}
  For a simply typed language $(\Sigma, \Omega)$, the family of sets of \emph{raw terms} indexed by a list~$V$ of variables consists of
  \begin{inlineenum}
    \item variables in $V$,
    \item annotations $\isTerm{t \annotate A}$ for some raw term $t$ in $V$ and a type $A$, and
    \item a construct $\tmOpts$ for some $\bop$ in~$O$, where $\vars{x}_{\isTerm{i}}$'s are lists of variables whose length is equal to the length of~$\Delta_i$, and $t_i$'s are raw terms in the variable list $V, \vars{x}_i$.
  \end{inlineenum}
  These correspond to rules $\Rule{Var}$, $\Rule{Anno}$, and $\Rule{Op}$ in \Cref{fig:raw-terms} respectively.
\end{definition}

@@

The definition of typing derivations is more involved.
We need some information to compare types on the object level during type synthesis and substitute those type variables in a typing derivation $\Gamma \vdash \tmOpts : A$ for an operation $o$ in $\Omega$ at some point.
Here we choose to include a substitution $\rho$ from the local context $\Xi$ to $\emptyset$ as part of its typing derivation explicitly:

@@

\begin{figure}
  \centering
  \small
  \judgbox{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^\dir{d} A}{$\isTerm{t}$ has a type $A$ in mode $\dir{d}$ under $\Gamma$ for a bidirectional system $(\Sigma, \Omega)$} 
  \begin{mathpar}
    \inferrule{(x : A) \in \Gamma}{\Gamma |-_{\Sigma, \Omega} \isTerm{x} :^{\syn} A}\,\SynRule{Var}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}{\Gamma |-_{\Sigma, \Omega} (\isTerm{t \annotate A}) :^{\syn}  A}\,\SynRule{Anno}
    \and
    \inferrule{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{{\syn}} B \\ B = A}{\Gamma |-_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}\,\ChkRule{Sub}
    \\
    \inferrule{\rho\colon \Sub{\Xi}{\emptyset} \\ \Gamma, \simsub{\vars{x}_\isTerm{1} : \Delta_{1}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_{1}} :^\dir{d_1} \simsub{A_1}{\rho} \\
      \cdots \\
    \Gamma, \simsub{\vars{x}_\isTerm{n} : \Delta_{n}}{\rho} |-_{\Sigma, \Omega} \isTerm{t_n} :^{\dir{d_n}} \simsub{A_{n}}{\rho}}
    {\Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}} \,\mathsf{Op}
    \\ \text{for $\biop$ in $\Omega$}
  \end{mathpar}
  \caption{Bidirectional typing derivations}
  \label{fig:bidirectional-typing-derivations}
\end{figure}

\begin{definition}\label{def:bidirectional-typing-derivations}\label{def:mode-derivations}
  For a bidirectional type system $(\Sigma, \Omega)$,
  \begin{itemize}
    \item the set of \emph{bidirectional typing derivations} of $\Gamma \vdash_{\Sigma, \Omega} t :^\dir{d} A$, indexed by a typing context $\Gamma$, a raw term $\isTerm{t}$ under $\erase{\Gamma}$, a mode $\dir{d}$, and a type $A$, is defined in \Cref{fig:bidirectional-typing-derivations} and particularly
          \[
            \Gamma |-_{\Sigma, \Omega} \tmOpts :^{\dir{d}} \simsub{A_0}{\rho}
          \]
          has a derivation for $\biop$ in $\Omega$ if there is $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$ and a derivation of $\Gamma, \vars{x}_{\isTerm{i}} : \simsub{\Delta_i}{\rho} \vdash_{\Sigma, \Omega} \isTerm{t_i} :^{\dir{d_i}} \simsub{A_i}{\rho}$ for each $i$;
    \item the set of \emph{mode derivations} of $V |-_{\Sigma, \Omega} t^\dir{d}$, indexed by a list $V$ of variables, a raw term $\isTerm{t}$ under $V$, and a mode $d$, is defined in \Cref{fig:mode-derivations}.
  \end{itemize}
  The two judgements {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \syn A}$} and {\small$\boxed{\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} \chk A}$} stand for ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\syn} A}$ and ${\Gamma \vdash_{\Sigma, \Omega} \isTerm{t} :^{\chk} A}$, respectively.
  A typing rule is \emph{checking} if its conclusion mode is $\chk$ or \emph{synthesising} otherwise.
\end{definition}

\begin{figure}[b]
  \centering
  \small
  \judgbox{V |-_{\Sigma, \Omega} \isTerm{t}^\dir{d}}{$\isTerm{t}$ is in mode $d$ with free variables in $V$ for $(\Sigma, \Omega)$}
  \begin{mathpar}
    \inferrule{x \in V}{V |-_{\Sigma, \Omega} \isTerm{x}^{\syn}}\,\SynRule{Var}
    \and
    \inferrule{\cdot |-_{\Sigma} A \\ V |-_{\Sigma, \Omega}\isTerm{t}^{\chk}}{V |-_{\Sigma, \Omega} (\isTerm{t \annotate A})^{\syn}}\,\SynRule{Anno}
    \and
    \inferrule{V |-_{\Sigma, \Omega} \isTerm{t}^{\syn}}{V |-_{\Sigma, \Omega} \isTerm{t}^{\chk}}\,\ChkRule{Sub}
    \and
    \inferrule{V, \vars{x}_1 |-_{\Sigma, \Omega} \isTerm{t_1}^\dir{d_1} \\ \cdots \\ V, \vars{x}_{n} |-_{\Sigma, \Omega} \isTerm{t_n}^\dir{d_n}}
    {V |-_{\Sigma, \Omega} \tmOpts^\dir{d}}\,\Rule{Op}
    \and \text{for $\biop$}
  \end{mathpar}
  \caption{Mode derivations}
  \label{fig:mode-derivations}
\end{figure}
Every bidirectional binding signature $\Omega$ gives rise to a binding signature $\erase{\Omega}$ if we erase modes from $\Omega$, called the \emph{(mode) erasure} of $\Omega$.
Hence a bidirectional type system $(\Sigma, \Omega)$ also specifies a simply typed language $(\Sigma, \erase{\Omega})$, including raw terms and typing derivations.
